---
title: "PM 566 Final Project"
author: "Erica Shin"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

## PM 566 Health Data Science: Final Project

**Pediatric Leukemia Cancer Cases (\<10 years) in the U.S., 2000-2021**

## Description of Project

### **Project Overview:**

This project investigates the demographic factors influencing the survival time of pediatric leukemia patients (under age 10) in the U.S. using data from the SEER (Surveillance, Epidemiology, and End Results) program, which collects cancer incidence data across a significant portion of the U.S. population. The focus of this analysis is to explore how sex, race, median income, and age at diagnosis are associated with survival outcomes for children diagnosed with leukemia.

### **Research Question:**

The primary question addressed in this project is:\
*What demographic factors (e.g., sex, race, median income, age) are associated with the likelihood of longer survival times in pediatric patients under age 10 diagnosed with malignant leukemia in the U.S.?*

The project seeks to identify if disparities in survival times exist based on sex, race, socioeconomic status, and age, contributing to a deeper understanding of how these factors influence the likelihood of longer survival and, ultimately, the development of targeted healthcare interventions for at-risk populations.

### **Data and Methods:**

The dataset used in this analysis is derived from the SEER program, specifically the "*Incidence - SEER Research Limited-Field Data, 22 Registries, Nov 2023 Sub (2000-2021) - Linked to County Attributes - Time Dependent (1990-2022) Income/Rurality, 1969-2022 Counties*" registry. The dataset covers a wide range of variables, including patient demographics, age, year of diagnosis, site of leukemia, year of death, and median household income. After cleaning and wrangling the data (removing missing values and transforming variables), the focus is on deceased patients to analyze survival time.

The following methods were applied:

-   **Data Cleaning and Transformation**: Renaming variables for simplicity, converting categorical variables to numeric where appropriate, and filtering out "alive" patients to focus on those who have passed away.

-   **Exploratory Analysis**: Using summary statistics and visualizations to investigate trends in survival time based on demographic factors.

-   **Survival Time Calculation**: Creating a new variable for survival time by subtracting the year of diagnosis from the year of death.

### **Key Insights and Visualizations:**

The project includes interactive visualizations to explore the relationship between demographic factors (sex, race, median income, and age) and survival time. Some key insights include:

-   Survival trends across different races and sexes.

-   The impact of median household income on survival.

-   Age-related trends in survival for pediatric leukemia patients.

### **Conclusion:**

This project seeks to explore the demographic disparities in survival outcomes among pediatric leukemia patients, focusing on key insights that could guide healthcare strategies and interventions aimed at improving survival rates in underserved populations.

## Interactive Visualizations

```{r, echo=FALSE}
# Read in dataset
leuk <- read.csv("/Users/ericashin/Downloads/Pediatric Malignant Leukemia Cancer Cases in the US (2000-2021, Ages Under 10 Only).csv")
```

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
# Load in libraries
library(dplyr)
library(ggplot2)
library(stringr)
library(knitr)
library(kableExtra) #extra styling for tables
library(paletteer) #color palette
library(plotly)

# Rename variable column names
leuk <- leuk |>
  rename(id = Patient.ID,
         sex = Sex,
         race = Race.recode..White..Black..Other.,
         eth = Race.ethnicity,
         age = Age.recode.with.single.ages.and.90.,
         age_range = Age.recode.with..1.year.olds,
         year_dx = Year.of.diagnosis,
         site = Site.recode.ICD.O.3.WHO.2008,
         level = Behavior.code.ICD.O.3,
         year_death = Year.of.death.recode,
         med_income = Median.household.income.inflation.adj.to.2022,
         rep_source = Type.of.Reporting.Source,
         loc_type = Rural.Urban.Continuum.Code)

leuk <- leuk |> filter(!is.na(med_income))

# Change age from character to numeric variable
leuk <- leuk |>
  mutate(age = as.integer(sub(" years", "", age)))

# Convert year_death into numeric variable
leuk$year_death <- as.numeric(as.character(leuk$year_death))

# Filter out NA (cases that are still alive) and only keeping cases with year of death
leuk <- leuk |>
  filter(!is.na(year_death) & !is.na(year_dx))

# Create new variable for survival time in years (called surv)
leuk <- leuk |>
  mutate(surv = year_death - year_dx)
```

**Visualization 1.** Interactive Barchart of Average Survival Time (Years) by Sex

```{r, echo=FALSE}
# Interactive Barchart of Average Survival Time (Years) by Sex
one <- leuk |>
  ggplot(mapping = aes(x = sex, y = surv, fill = sex)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  labs(title = "Barchart of Average Survival Time (Years) by Sex",
       x = "Sex",
       y = "Average Survival Time (Years)",
       fill = "Sex") +
  theme_minimal() +
  scale_fill_paletteer_d("lisa::BridgetRiley", labels = function(x) str_wrap(x, width = 5)) +
  theme(legend.position = "none") # Hide the legend

# Convert the ggplot to an interactive plot using plotly
plot1 <- ggplotly(one)

# Show the interactive plot
plot1
```

Males had a higher average survival time (2.494 years) compared to females (2.168 years).

------------------------------------------------------------------------

**Visualization 2.** Interactive Barchart of Average Survival Time (Years) by Race

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Interactive Barchart of Average Survival Time (Years) by Race
two <- leuk |>
  ggplot(mapping = aes(x = race, y = surv, fill = race)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  labs(title = "Barchart of Average Survival Time (Years) by Race",
       x = "Race",
       y = "Average Survival Time (Years)",
       fill = "Race") +
  theme_minimal() +
  scale_fill_paletteer_d("lisa::BridgetRiley") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  theme(legend.position = "none") # Hide the legend

# Convert the ggplot to an interactive plot using plotly
plot2 <- ggplotly(two)

# Show the interactive plot
plot2
```

Among racial categories, the "Unknown" group had the highest average survival time at 2.450 years, closely followed by "Other (American Indian/AK Native, Asian/Pacific Islander)" at 2.446 years.

------------------------------------------------------------------------

**Visualization 3.** Interactive Barchart of Patient Count/Sample Size by Race

```{r, echo=FALSE, warning=FALSE}
# Interactive Barchart of Patient Count/Sample Size by Race using plotly
sum_race <- leuk |>
  group_by(race) |>
  summarize(
    Count = n(),
    Mean = mean(surv, na.rm = TRUE),
    Median = median(surv, na.rm = TRUE),
    SD = sd(surv, na.rm = TRUE),
    .groups = 'drop'
  )

plot3 <- plot_ly(sum_race, 
               x = ~race, 
               y = ~Count, 
               type = 'bar',  
               hoverinfo = ~paste("<br>Patient Count: ", Count),  
               marker = list(
                     color = paletteer_d("lisa::BridgetRiley")[factor(sum_race$race)]
                     )) |>
  layout(
    title = 'Barchart of Patient Count/Sample Size by Race',
    xaxis = list(
      title = 'Race',
      tickangle = 0,  
      tickmode = 'array',  
      tickvals = sum_race$race,  
      ticktext = sapply(sum_race$race, function(x) str_wrap(x, width = 10))
    ),
    yaxis = list(title = 'Patient Count')
  )

plot3
```

The majority of patients in the dataset were classified as "White" (2703 out of 3457 patients or 78%) while the "Unknown" category contained only 20 patients (5.8%), compared to 307 for "Other" (8.9%) and 427 for "Black" (12.4%).

------------------------------------------------------------------------

**Visualization 4.** Interactive Scatterplot of Average Survival Time (Years) by Median Income

```{r, echo=FALSE}
# Median Income

# Define a specific order for the income ranges
income_levels <- c("< $40,000", "$40,000 - $44,999", "$45,000 - $49,999", "$50,000 - $54,999", 
                   "$55,000 - $59,999", "$60,000 - $64,999", "$65,000 - $69,999", "$70,000 -   $74,999", "$75,000 - $79,999", "$80,000 - $84,999", "$85,000 - $89,999", "$90,000 - $94,999", "$95,000 - $99,999", "$100,000 - $109,999", "$110,000 - $119,999", "$120,000+")

# Convert med_income to a factor with defined levels
leuk <- leuk |>
  mutate(med_income = factor(med_income, levels = income_levels, ordered = TRUE))

# Summary statistics for survival times by median income, sorted by factor levels
sum_med_income <- leuk |>
  group_by(med_income) |>
  summarize(
    Count = n(),
    Mean = mean(surv, na.rm = TRUE),
    Median = median(surv, na.rm = TRUE),
    SD = sd(surv, na.rm = TRUE),
    .groups = 'drop'
  ) |>
  arrange(desc(med_income))  #sorting by the ordered factor

# Create the ggplot scatter plot
four <- sum_med_income |>
  filter(!is.na(med_income)) |>
  ggplot(mapping=aes(x=med_income, y=Mean)) +
  geom_point() +
  labs(title = "Scatter Plot of Median Income by Average Survival Time (Years)",
       x = "Median Income",
       y = "Average Survival Time (Years)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))

# Convert the ggplot to an interactive plot using plotly
plot4 <- ggplotly(four)

# Show the interactive plot
plot4
```

There appears to be a positive association between median income and average survival time, indicating that higher income is correlated with better survival outcomes.

------------------------------------------------------------------------

**Visualization 5.** Interactive Line Plot of Average Survival Time by Age of Diagnosis and Race

```{r, echo=FALSE}
# Interactive Line Plot of Average Survival Time by Age of Diagnosis and Race
avg_surv_age_race <- leuk |> 
  group_by(age, race) |> 
  summarize(avg_surv = mean(surv, na.rm = TRUE), .groups = 'drop')

five <- ggplot(avg_surv_age_race, aes(x = age, y = avg_surv, color = race, group = race)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Average Survival Time by Age of Diagnosis and Race",
       x = "Age of Diagnosis (Years)",
       y = "Average Survival Time (Years)",
       color = "Race") +
  theme_minimal() +
  scale_color_paletteer_d("lisa::BridgetRiley", labels = function(x) str_wrap(x, width = 5)) # wrapped labels

# Convert the ggplot to an interactive plot using plotly
plot5 <- ggplotly(five)

# Display the interactive plot
plot5
```

The discrepancy in sample sizes suggests that average survival times for the smaller race categories may be more susceptible to outliers, especially in the "Unknown" group, which exhibited an unusually high average survival time of around 15 years for patients approximately 2.5 years old.

------------------------------------------------------------------------

## Full Report

Download the PDF version of the full report [here](./docs/Final-Project-Report.pdf).
